A Real-Time Sonar Simulator for Automatic Target Recognition

### 2018-02-06 LCA/article.tex
Major commit before messing with PhD thesis

% \usepackage{fixltx2e}                           % Fix some twocolumn float problems

% \usepackage{stfloats}                          % Allows: \begin{figure*}[!b]
                                                % (double column figures on top/bottom)

@@

% NOTE: PDF hyperlink and bookmark features are not required in IEEE
%       papers and their use requires extra complexity and work.
% \newcommand\MYhyperrefoptions{bookmarks=true,bookmarksnumbered=true,
% pdfpagemode={UseOutlines},plainpages=false,pdfpagelabels=true,
% colorlinks=true,linkcolor={black},citecolor={black},urlcolor={black},
% pdftitle={Optimal Window Design for the Low Complexity Adaptive Beamformer in Active Sonar Imaging},
% pdfsubject={},
% pdfauthor={Jo Inge Buskenes},
% pdfkeywords={adaptive beamforming, beamforming, complexity, sonar, active}}%
\ifCLASSINFOpdf
%    \usepackage[\MYhyperrefoptions,pdftex]{hyperref}
   \usepackage[pdftex,
    final,                %Keep hyperlink stuff even when in draft mode
    breaklinks=true,      %Break links when necessary
    linktocpage=true,     %Enable link to page?
    linkcolor=black,  %Colour of links to labels within document
    citecolor=black,      %Colour of links to the biliography
    filecolor=red,        %Colour of links to local files
    pagecolor=black,        %Colour of links to other pages withing document
    urlcolor=Blue,        %Colour of the links to external URLs
    colorlinks=true,      %??
    plainpages=false,     %Store roman/arabic numbering differently to avoid
    bookmarksnumbered ]{hyperref}
% 
% \else
%    \usepackage[\MYhyperrefoptions,breaklinks=true,dvips]{hyperref}%
\fi
%    \usepackage{breakurl}                        % Allows 'dvips' driver to break links

@@

\let\oldincludegraphics\includegraphics
\renewcommand\includegraphics[2][]{%
  \immediate\write18{./bin/laFigure #2 #1}%
  \input{result}}%

@@

\newcommand\Om{O_\text{m}}
\newcommand\Oa{O_\text{a}}
\newcommand\Nl{N_\text{l}}
\newcommand\Nk{N_\text{k}}
\newcommand\1{\vec 1}
\newcommand\I{\mat I}
\renewcommand*\a{\vec a}
\newcommand*\f{\vec f}
\renewcommand*\i{\vec i}
\renewcommand*\k{\vec k}
\newcommand*\n{\vec n}
\newcommand*\p{\vec p}
\newcommand*\s{\vec s}
\newcommand*\w{\vec w}
\newcommand*\x{\vec x}
\newcommand*\y{\vec y}

\newcommand*\A{\mat A}
\newcommand*\B{\mat B}
% \newcommand*\C{\mat C}
\newcommand*\E{\mat E}
\renewcommand*\P{\mat P}
\newcommand*\eP{\mat{\hat P}}
\newcommand*\R{\mat R}
\newcommand*\Ri{\R^{-1}}
\newcommand*\eR{\mat{\hat R}}
\newcommand*\eRi{\hat{\mat R}\;\!^{-1}}
\newcommand*\Navg{N_\text{avg}}
\newcommand*\W{\mat W}
\newcommand*\X{\mat X}
\newcommand*\Xd{\X_{\!\Delta}}
\newcommand*\Y{\mat Y}

\renewcommand\argmin{\text{argmin}}

@@

\renewcommand*\L{\mat \Lambda}
% \newcommand*\U{\mat U}
% \renewcommand*\t{\mathtiny{^T}}
% \newcommand*\h{\mathtiny{^H}}
\renewcommand*\t{^T}
\newcommand*\h{^H}

\newcommand\minus{\scalebox{0.75}[1.0]{$-$}}

% \usepackage{tikz}
% \usetikzlibrary{shapes,snakes}
\usepackage{amsmath,amssymb}
% \usepackage{datatool}
% \usepackage{glossaries}

% \newenvironment{outline}
% {\begin{itemize}}
% {\end{itemize}}

@@

@@

\setcounter{totalnumber}{1}
\setcounter{topnumber}{1}
\setcounter{dbltopnumber}{1}

\begin{figure*}[t]\centering%
\includegraphics[width=\linewidth]{gfx/scene_coordinate_system.svg}%
\caption{Simulator principle. A model of the seafloor and object is first loaded into OpenGL, and an orthogonal camera view set up to match the desired image region on the seafloor. Then the scene is rendered and an intensity image and a depth map is created. The intensity image along with the depth information is finally combined to form a sonar like image. This is our template.}\label{simulator_coordinate_system}%
\end{figure*}

\begin{figure*}[t]\centering%
\includegraphics[drawing,width=\linewidth]{gfx/simulator.svg}%
\caption{Simulator concept.}\label{buildup}%
\end{figure*}

% \IEEEPARstart{W}{hen} 
% Autonomous underwater vehicles (AUVs). When an AUV with a sonar imaging system encounter an interesting object, it would be very beneficial if adapted its behavior to examine the object further. This is challenging as it would need to learn in what ways the interesting objects differ from the background, i.e. it relies on previous knowledge. It would also need to process this information is near real-time to act in an efficient manner.

% - Adapt simulation to its behaviour, see if they still match.

\IEEEPARstart{A}{utonomous} underwater vehicles (AUVs) are becoming increasingly important in human underwater endeavors. When equipped with a synthetic aperture sonar (SAS), they can e.g. be used to map the seafloor, survey underwater biology, help localize wrecks, inspect oil pipes, fields and electricity lines, or hunt for mines and biohazard waste. Such independent systems have much to gain from being able to adapt to the scenario at hand. For instance, with a well designed automatic target recognition (ATR) system on board they can adapt their mission plan to collect more information on particularly interesting objects, or fine-tune the sonar and navigation system to improve imaging performance.

@@

In conventional ATR an object of interest is first isolated, its properties attempted estimated, and a classifier used to determine what object class these properties belong to. The challenge is finding unique and intrinsic object properties, such as its type, shape and size, and to do so independently of extrinsic properties such as its alignment, burial depth, degeneration level or view angle. While SAS provide images with state of the art resolution and dynamic range, they still only represent an ambiguous acoustical 2D projection of the 3D object, littered with a highly coherent noise floor.

An alternative approach is to use template matching, where the isolated image segment is compared to a set of template images covering the relevant object classes, and assigned to the class of the template with the best fit. Since the templates are computed from 3D models, this reduces the issue of image ambiguity. However, a challenge is finding a set of templates that represent the actual configurations of object and seafloor, Another issue is that the size of precomputed library of templates grows exponentially with the number of parameters needed. This limits its use to only a few parameters being coarsely sampled, ultimately yielding inaccurate results~\cite{Midelfart2010}.

% We avoid the problems of conventional template matching by computing the templates on-the-fly as part of the classification process, thereby not needing a precomputed set of templates at all. 
%  that avoids this problem by running sufficiently fast to create templates adapted to the actual scene as part of the classification process.
% Classification process

We tackle the problems of conventional template matching by performing:
%
\begin{itemize}
\item \emph{Adaptive template matching}; where a template is selected based on object and seafloor parameters estimated from the SAS image~\cite{Midelfart2010}, and
\item \emph{Real-time template simulation}; where a template is created tailored to the estimated parameters, improving its accuracy and avoiding the need of a precomputed template library.
\end{itemize}
%

The methods presented herein will focus on the template simulator, but some details will also be given on adaptive template matching for completeness. More details on this technique can be found in~\cite{Midelfart2010}.

 This is achieved with the aid of the massive computing power in graphics processing units (GPUs) and optimized software libraries for scene rendering. The simulator loads a 3D model of the seafloor and object class into OpenGL and positions the camera at the sonar location. The scene is highlighted using a light tube that extends to infinity along the AUV's axis of probagation. We assume rough, isotropic surfaces reflecting sound energy equally in all directions. This can be modeled with a Lambertian scattering model~\cite{Blake1993,Bell1995}, where the intensity of the backscattered sound solely depends on the incidence angle of the transmitted signal onto the model surface. The OpenGL rendering produces an optical 2D intensity image and depth map that reveals the distance from each pixel to the propagation axis. For maximum flexibility this data is finally combined with OpenCL to produce the image templates.

@@

For lack of a better name, the simulator will be referred to as "FFISim" throughout this article.

Show:
\begin{itemize}
\item Adaptive template matching to avoid large static template libraries,
\item Improved classification performance by feeding simulator with parameters derived off SAS image,
\item Further improved classification by performing a fine-search within a narrow search space around the SAS derived parameters. GPU makes this more powerful.
\end{itemize}

@@

Models are not ROCCAN (???)

\subsection{Loading the scene}

Our models are stored in regular 3D model files. The simulator begins by loading these into a tree of nodes, each representing a vertex in the model with corresponding facet and texture information. Then the vertices are placed into a virtual scene in OpenGL, and a camera set up to capture the part of the scene corresponding to the image. When OpenGL renders the image, it first computes which vertex contributes to which pixel in the image, then computes the intensity and color of the pixel based on the node's facet, texture and scattering model. The code that operates on vertices are called a vertex shader, and the code that operates on pixels are called a fragment shader.

The mapping of the vertices from object coordinates to image coordinates happen through a set of transformations:
%
\begin{align*}
\bmat{x\\y\\0\\1}_\text{image} = \boldsymbol{P} \cdot \boldsymbol{V} \cdot \boldsymbol{M} \cdot \bmat{x\\y\\z\\1}_\text{object},
\end{align*}
%
where $\bmat{\cdot}_\text{model}$ is the loaded model, $\bmat{\cdot}_\text{image}$ is the output image, and $\M$, $\V$ and $\P$ are transformation matrices. These will be described next.

The first transform we apply is the model matrix $\M$, which operates in the model's local coordinate system. It is defined as
%
\begin{align}
\M = \bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T} \cdot \boldsymbol{R} \cdot \boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model},
\end{align}
%
where $\boldsymbol{S}$, $\boldsymbol{R}$ and $\boldsymbol{T}$ are matrices\todo{be explicit on dependencies?} for scaling, rotation and translation, respectively. These are provided for reference in Appendix \ref{transformation_matrices}.

Next we map the model into world coordinates using the view matrix $\V$. This involves aligning the camera to precisely fit the image region. In our case the camera represents the phase center of the sonar, and it is placed in the world origin. The transformation then take the form of
%
\begin{align}
\V &= \mat{T}(x_\text{c}, y_\text{c}, z_\text{c}) \cdot \R_\text{x}(\phi_x, 0, 0) \nonumber\\
&= 
\bmat{
1  &  0  &  0  &  x_\text{c} \\
0  &  1  &  0  &  y_\text{c} \\
0  &  0  &  1  &  z_\text{c} \\
0  &  0  &  0  &  1 \\
} \cdot
\bmat{
1  &  0           &  0           &  0 \\
0  &  \cos\phi_x  &  -\sin\phi_x &  0 \\
0  &  \sin\phi_x  &  \cos\phi_x  &  0 \\
0  &  0           &  0           &  1 \\
},
\end{align}
%
where $(x_\text{c}, y_\text{c}, z_\text{c})$ is the image center coordinates for along track, cross track, and depth, respectively, and $\phi_x$ is the elevation angle\todo{is this correct?} of the sonar. It is given as
%
\begin{align*}
\phi_x &= \arctan\left(\frac{y_\text{mean}}{-z_\text{mean}}\right).
\end{align*}\todo{use absolute sign?}

Finally the relevant part of the scene is projected onto the image plane with a projection matrix $\P$. There are two main types of projection, perspective and orthographic. Perspective projection scales all $x$ and $y$ coordinates inversely with the distance to the objects to create a sense of perspective in the image. Orthographic projection, on the other hand, maps all the scene's vertices to a squared cube where each dimension has the range [-1,1]. This will render all objects equally large whether they are in the distance or up close. The matrix that gives orthographic projection is defined as
%
\begin{align}
\P = \left(\begin{matrix}
\frac{2}{r-l}  &  0              &  0              &  -\frac{r+l}{r-l} \\
0              &  \frac{2}{t-b}  &  0              &  -\frac{t+b}{t-b} \\
0              &  0              &  \frac{-2}{f-n} &  -\frac{f+n}{f-n} \\
0              &  0              &  0              &  1
\end{matrix}\right)
\end{align}
%
where $l,r,b$ and $t$ are the left, right, bottom and top boundaries of the projection plane, respectively, and where $n$ and $f$ is the near and far limits of the cube that encompass the scene. These parameters need to be adapted to provide a 2D image that matches the image size, which we specify in along track and cross track coordinates ($x,y$). This is straightforward for left and right since matches the $x$-axis
% They define the window through which we observe our scene.
\begin{align}
l &= x_\text{min}  &  r &= x_\text{max}
\end{align}
%
For the near and far limits one simply needs to make sure the scene will fit in it, say
%
\begin{align}
n &= 0\;\text{m}   &  f &= 300\;\text{m}.
\end{align}
%
The bottom and top boundaries can be inferred from \Fig{camera_top_bottom}. First we define an origin vector that points to the center of the image
%
\begin{align}
\vec o &= \left[\frac{x_\text{min}+x_\text{max}}{2}, \frac{y_\text{min}+y_\text{max}}{2}, \frac{z_\text{min}+z_\text{max}}{2} \right]\T,
\end{align}
%
and the distance from the image origin to its $y$-boundaries is given as
%
\begin{align}
\Delta y &= \frac{|y_\text{max} - y_\text{min}|}{2}.
\end{align}
%
To find the nearest and farthest point in the image that will be part of the projected scene, $\alpha$ and $\beta$, we must consider the seafloor rotation angle around the $x$-axis, $\Phi_x$ 
%
\begin{align}
\vec\alpha &= \vec 0 - \left[0, \Delta y, \Delta y \cdot \tan\Phi_\text{x} \right] \nn
\vec\beta &= \vec 0 + \left[0, \Delta y, \Delta y \cdot \tan\Phi_\text{x} \right]
\end{align}

\begin{figure}[t]\centering%
\includegraphics[drawing,width=\linewidth]{gfx/opengl_part.svg}%
\caption{Setting camera top and bottom boundaries.}\label{camera_top_bottom}%
\end{figure}

@@

\todo{
- Kernel tuning.
- Sharing. PBO. Asynchronous.
}

\subsection{Template matching}

1. Segmentation
2. Parameter estimation
   Available: Range, altitude, resolution => Input to both standard and adaptive. In standard the best fit out of a predefined template database is selected, then resampled to correct resolution.
   Additional for adaptive, parameters that modify object model: Target aspect angle, seafloor slope, object length and height.

   Target aspect angle computed with Radon Transform, seafloor under object modeled as plane with a slope estimated from bathymetric data, length estimated from longest part of highlight region, and height estimated from the shape of the shadow. The burial depth is n

3.

   Details: \ref{Midelfart2010}.

Let $S(x,y)$ be the isolated part of the

Matching the isolated image segment with the template take the form of

\begin{align}
c_{h,i} = \underset{x',y'}{\text{argmax}} \sum\limits_{\forall x}\sum\limits_{\forall y} \big| H(x,y) - T(x-x',y-y') \big|
\end{align}

\begin{align}
c_i = \frac{c_{h,i} + c_{s,i}}{2}
\end{align}

And the final template used is the one with the greatest correlation coefficient

\begin{align}
\underset{T_i(x,y)}{\text{argmax}} C_i\sum\limits_{\forall x}\sum\limits_{\forall y}  S(x,y) T(x-x',y-y')
\end{align}

Then assign object to class $i$.

To narrow down the search space of possible templates we derive some reliable statistics of the seafloor and object from the SAS image. While FFISim accepts an arbitrarily complex seabed, we find that modeling it as a plane wave with the right tilt works well. The object is more difficult. For instance, we estimate its length and height from the shape of its shadow, but this fails if the object is notably immersed into the seabed at one end.

After the a template is created we decompose it into its highlight and shadow component, and compute a correlation score between these and the corresponding areas in the SAS image. Then the two scores are summed and normalized to form the final correlation score of the template.

We tag the template matchers as either semi- or fully adaptive. The former refers to using the best matching SIGMAS+ generated template out of a precomputed set of templates, while the latter refers to the proposed method of using FFISim to generate templates on the-the-fly. \todo{Should be more precise. How many templates? Parameter span? How finely were the parameters sampled?}

@@

\todo{
Write to file
}

%    """
%    
%    stm: Standard template matching, made with SIGMAS.
%    atm: Adaptive template matching, made with FFISim.
%    atm_bury: Adaptive template matching where a search of bury depths is used.
%    atm_bury_restr: Restricted?
%    
%    Adaptiveness made by letting the template matcher use 5 different rotation
%    angles in the ground-range axis. 
%    
%    
%    Data from Jesusbukta
%    
%    Figures are from run 2012.06.16
%    det210 - Cylinder 1: est 2.92m (short side towards sonar)
%    det233 - Cylinder 2: est 3.05m (diagonal)
%    det379 - Torpedo: est 5.06m (looks good)
%    
%    ROC curves from 3 different runs, first two just 2 days apart in 2009, last one in 2012.
%    
%    run_20120616 - last one. Much better with adaptive templates.
%    
%    ROC curves made by average correlation score of shadow/highlight from template matching.
%    
%    
%    fig4b(2015.11.16)
%       Bug that skipped masking of port side (negative y) detections past 180m range fixed
%       
%    fig4 (2015.11.16)
%       Bug leading to incorrect estimation of cylinder orientation fixed.
%          This affects run090601_1. Should be better after this.
%          
%    fig3 (2015.11.09)
%       Detections outside 180m range have been masked (but bug for port side fixed in fig4b).
%       Fixed an incorrectly annotated detection.
%       
%    fig2 (2015.11.02)
%       Height of object estimated from shadow length
%       Let adaptive method try 5 different rotations in ground-range (-4, -2, 0, 2, 4)
%          Improved results, but incorrect classification of:
%          5 detections in first run, and
%          3 detections in second run
%          => For FPR=0.2 the results are slightly worse than with normal adaptive (particularly first run).
%          Images are problematic in these cases:
%             2 detections get wrong orientation because a line (anchor) cross the image
%             Other misclassifications in image regions with poor echo or shadow strength
%          => Not a big issue as not enough time to detect more than ~20% of the detections in a run
%             In a MCM operation these detections would never be used.
%             The runs have 3000 detections each, photographing 20% of these is too much.
%          Performance FPR>0.2 not interesting.
%          Performance FPR<0.2 much better with adaptive template matching.
%          
%    fig (2015.10.23)
%       atm - cylinder orientation and length estimated from SAS image and fed into simulator.
%       
%       ROC curves from last run. Computing the mean correlation score of the shadow and highlight/echo.
%       Early results. Can only check vs. cylinder-like objects.
%       Standard template library created based on estimation of physical size of known cylinder mines.
%       Not possible for adaptive method to perform better when hitting such a target.
%       
%  
%    """

>>>>

\usepackage{fixltx2e}                           % Fix some twocolumn float problems

\usepackage{stfloats}                          % Allows: \begin{figure*}[!b]
                                                % (double column figures on top/bottom)

@@

% NOTE: PDF hyperlink and bookmark features are not required in IEEE
%       papers and their use requires extra complexity and work.
\newcommand\MYhyperrefoptions{bookmarks=true,bookmarksnumbered=true,
pdfpagemode={UseOutlines},plainpages=false,pdfpagelabels=true,
colorlinks=true,linkcolor={black},citecolor={black},urlcolor={black},
pdftitle={A Fast GPU Sonar Simulator for Automatic Target Recognition},
pdfsubject={},
pdfauthor={Jo Inge Buskenes},
pdfkeywords={adaptive beamforming, beamforming, complexity, sonar, active}}%
\ifCLASSINFOpdf
   \usepackage[\MYhyperrefoptions,pdftex]{hyperref}
\else
   \usepackage[\MYhyperrefoptions,breaklinks=true,dvips]{hyperref}
   \usepackage{breakurl}                        % Allows 'dvips' driver to break links
\fi

@@

\newcommand\graphicsAI[2][]{%
  \immediate\write18{./bin/laFigure #2 #1}%
  \input{result}}%

@@

\newcommand\Om{O_\text{m}}
\newcommand\Oa{O_\text{a}}
\newcommand\Nl{N_\text{l}}
\newcommand\Nk{N_\text{k}}
\newcommand\1{\vec 1}
\newcommand\I{\mat I}
\renewcommand*\a{\vec a}
\renewcommand*\i{\vec i}
\renewcommand*\k{\vec k}
\newcommand*\n{\vec n}
\newcommand*\p{\vec p}
\newcommand*\s{\vec s}
\newcommand*\w{\vec w}
\newcommand*\x{\vec x}
\newcommand*\y{\vec y}

\newcommand*\A{\mat A}
\newcommand*\B{\mat B}
\newcommand*\C{\mat C}
\newcommand*\E{\mat E}
\renewcommand*\P{\mat P}
\newcommand*\eP{\mat{\hat P}}
\newcommand*\R{\mat R}
\newcommand*\Ri{\R^{-1}}
\newcommand*\eR{\mat{\hat R}}
\newcommand*\eRi{\hat{\mat R}\;\!^{-1}}
\newcommand*\Navg{N_\text{avg}}
\newcommand*\W{\mat W}
\newcommand*\X{\mat X}
\newcommand*\Xd{\X_{\!\Delta}}
\newcommand*\Y{\mat Y}

@@

\renewcommand*\L{\mat \Lambda}
\newcommand*\U{\mat U}
% \renewcommand*\t{\mathtiny{^T}}
% \newcommand*\h{\mathtiny{^H}}
\renewcommand*\t{^T}
\newcommand*\h{^H}

\usepackage{tikz}
\usetikzlibrary{shapes,snakes}
\usepackage{amsmath,amssymb}
% \usepackage{datatool}
% \usepackage{glossaries}

\newenvironment{outline}
{\begin{itemize}}
{\end{itemize}}

% correct bad hyphenation here
% \hyphenation{op-tical net-works semi-conduc-tor}

%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% GLOSSARY %
%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

% \makeglossaries
% \input{../../Library/glossary}

%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% DOCUMENT START %
%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

% \usepackage{yfonts}

% \graphicspath{submission/final/gfx/}

@@

\maketitle

@@

One of the major shortcomings of autonomous underwater vehicles (AUVs) is their lack of curiosity. When encountering an interesting object, an AUV with a mounted sonar imaging system should be able to detect it, and perhaps move in for closer examination. Besides the apparent challenge of trying to detect something with a long list of unknown parameters, this must happen in real-time and be very power efficient.

@@

The detection process is well described in the field of automatic target recognition (ATR). The classical approach involves isolating an object of interest, extracting parameters from it and feeding those into a classifier. However, it is hard to find parameters that describe the objects uniquely, in particular due to the ambiguities arising from interpreting a 3D world from its 2D projection. Another approach is to use template matching, where the image segment is compared to a set of template images covering the relevant object classes, and assigned to the class of the template yielding the the best fit. The challenge here is to come up with accurate template images covering the actual configurations of object and seabed, especially if a static template library is used, which will lead to a computational complexity that is exponential with the number of model parameters. This limits its use to only a few parameters being coarsely sampled, ultimately yielding inaccurate results~\cite{Midelfart2010}.

We have developed a synthetic aperture sonar (SAS) simulator that avoids this problem by running sufficiently fast to create templates adapted to the actual scene as part of the classification process. This is achieved with the aid of the massive computing power in graphics processing units (GPUs) and optimized software libraries for scene rendering. The simulator loads a 3D model of the seafloor and an object class model into OpenGL, where emitted sound waves are modeled with a light tube extending all the way down the propagation path of the AUV. We assume rough, isotropic surfaces reflecting sound energy equally in all directions. This can be modeled with a Lambertian scattering model~\cite{Blake1993,Bell1995}, where the intensity of the backscattered sound solely depends on the incidence angle of the transmitted signal onto the model surface. When rendering OpenGL is set up to produce an optical 2D intensity image and depth map that reveals the distance from each pixel to the propagation axis. For maximum flexibility this data is finally combined with OpenCL to produce the image templates.

The alternative of using of a static, predetermined template library has a computational complexity that is exponential with the number of parameters. This limits its use to only a few parameters being coarsely sampled, ultimately yielding inaccurate results~\cite{Midelfart2010}.

To avoid this problem we have developed a very fast side-looking sonar simulator that generates templates adapted to the actual scene.

synthetic aperture sonar (SAS)

@@

\begin{itemize}
\item SWAT (Shallow Water Acoustic Toolset): 
\item Bell's thesis:
\item SIGMA:
\item OpenGL does not support scattered writes. OpenCL does.
\item 2D template approximation:
\item Geometric distortions in SSS \cite{Cobra1992} Bell
\item Sidescan sonar and synthetic aperture sonar. In seafloor coordinates. Maps to orthogonal rendering.
\item Layover causes ambiguity, different seafloor phenomena can cause the same output image. Lots of cornercases when approximating the model in 2D. Instead use a 3D model and project it to a 2D one. Easier to operate, models are more general purpose. Less parameters to adjust.
\item Viewer can be used to train an operator for finding non-trivial objects on the seafloor. Input for automated target recognition standard.
\item Improved image quality of selected object, reduced data collection cost (price estimate?).
\item OpenGL rendering can have infinite resolution, only limited by the pixel density in the image.
\item Accuracy in terms of classification result. Need to embody static effects in a deterministic way. Incoherent synthesis of the image.
\item Sound speed assumed constant.
\item High frequency lead to little (define) penetration, lower range and higher resolution. Often modeled accurately with ray tracing techniques.
\end{itemize}

Model
\begin{itemize}
\item Ray solution: Separate amplitude and phase components of the Helmholtz eq. by applying geometric optics approximation. Ok as long as amplitude varies more slowly with position than the phase. Implications: High frequency, because the curvature of a ray over a wavelength must be small, and the fractional change in the sound speed must be small over a wavelength. Further simplification available by Snell's Law. Fast to compute. Not able to handle diffraction, focal points and caustics, and will cause (too) abrupt changes on distinct changes in sea environment. Can be improved upon with fuzzy beam modeling / Gaussian beam tracing.
\begin{itemize}
\item Pro: Easy to compute. Visual result. Handles complex bathymetry. Array and model response can be accounted for.
\item Con: Only works for high frequency and small sound variations. No diffraction, caustics and focus. Infinite energy caustics/sharp variations. Incoherent(?)
\end{itemize}
\item Normal modes: Separate amplitude and phase into a range dependent term, and a depth dependent term. Assumes minimal range dependence in sound speed profile and depth. Limited near field accuracy. Acoustic field does not need to be calculated at intermediate ranges. Penetration can be modeled. can be calculated 
\begin{itemize}
\item Pro: Ok to compute. Handles penetration. Low frequency, shallow water.
\item Con: Horizontal stratification. Far field only. 
\end{itemize}
\item Parabolic equation: One way propagation. Marching solution. Initial starting solution must be known, often computed with normal mode. Only valid for narrow beams $\pm$20. Able to handle range dependent environments. Computing full field is easy. Computation time is of O(f$^2$), hence mostly useful for low-frequency scenarios. Unable to deal with horizontal refraction (shadow zones from e.g. conical mounts)
\item Fast field programs / Green's function solutions. Exact full wave solution for acoustic fields in horizontally stratified media. Can handle both compressional and shear waves, and hence both fluid and solid media. Far field assumption. Similar to normal mode techniques.
\item Finite element: Scene segmented into triangles or rectangles with a size of one tenth of a wavelength, or less. At each node the wavefield can be determined from the wave equation. Computationally limited to low frequencies.
\item 2D model: Acoustic field in range and depth. Assumes that a wave transmitted in a vertical plane remains in it. Reasonable when the sound speed can be assumed constant in the horizontal plane, and when the seafloor normals are in the vertical range-depth plane (avoid horizontal refraction, or actually reflection).
\end{itemize}

What's new?
\begin{itemize}
\item Tuned for speed using GPUs, combining OpenGL and OpenCL to take advantage of existing functionality for computer graphics with the added flexibility of general purpose programming on the GPUs.
\item Performance to spare: Game-like viewer for the visualization and verification of the rendered scene and corresponding sonar image.
\item First simulator of its kinds aimed towards improved ATR and ultimately better AUVs.
\item 3D models are still novel. 
\end{itemize}

% The ATR algorithms must, however, be sufficiently fast to process the sensor data onboard the AUV in delayed real-time.

% We have developed a simulator for synthetic aperture sonar (SAS) that creates object templates very quickly with the aid of the massive computing power in graphics processing units (GPUs). It loads a 3D model of the sea floor and an object of interest into OpenGL, where emitted sound waves are modeled with a light source placed at the sonar transmit location. We assume rough, isotropic surfaces that reflects sound energy equally in all directions. This can be modeled with a Lambertian scattering model~\cite{Bell1997} where the intensity of the backscattered sound is said to depend only on the on incidence angle of the transmitted signal onto the model surface. When rendering OpenGL is set up to produce an optical 2D intensity image and depth map that reveals the distance from each pixels to the propagation axis. For maximum flexibility this data is finally combined with OpenCL to produce the image templates.
% 
% By coincidence our implementation is not so different from the SIGMAS+ simulator developed at the NATO Undersea Research Centre (NURC)~\cite{Coiras2009a, Coiras2009b}. They both take advantage of OpenGL to render a 2D image from 3D models, are designed for side-looking sonar, and assume a Lambertian scattering model. However, while SIGMAS+ is geared more towards realistically looking sonar images, and include effects such as noise, we do not. Also, their simulator sums images obtained from multiple OpenGL rendering passes to create the sonar image, while we render once and post-process this result with OpenCL to tailor the sonar image full flexibility.

\begin{figure}[t]\centering%
\graphicsAI[drawing,width=\linewidth]{gfx/specs.svg}%
\caption{Simulator specifications.}\label{specs}%
\end{figure}

\begin{figure*}[t]\centering%
\graphicsAI[drawing,width=\linewidth]{gfx/simulator.svg}%
\caption{Simulator concept.}\label{buildup}%
\end{figure*}

@@

\subsection{Setting up the scene}

Before we can render anything with OpenGL we first have to set up the scene. This involves loading the models and then placing them where we want them with the proper orientation.

Our models are stored in regular 3D model files. The simulator loads these with the Open Asset Import library (\texttt{assimp}). This is a portable open source library that supports loading a wide range of 3D model formats in a uniform manner. It converts the model into a node tree where each node contains data such as vertices, facets and textures, all formatted in an OpenGL friendly way.

The next step is to scale, orient and position the objects in the scene, and finally project this scene onto an image plane. This is achieved by applying a set of transformations to each vertex in the model. In OpenGL, the set of transformations can be formalized as
\begin{align*}
\bmat{x\\y\\0\\1}_\text{image} = \P \cdot \V \cdot \M \cdot \bmat{x\\y\\z\\1}_\text{model},
\end{align*}
where $\bmat{\cdot}_\text{model}$ is the loaded model, $\bmat{\cdot}_\text{image}$ is the output image, and $\M$, $\V$ and $\P$ are transformation matrices. These will be described next.

The first transform we apply is the model matrix $\M$, which operates in the model's local coordinate system. It is defined as
\begin{align}
\bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T} \cdot \R \cdot \boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model},
\end{align}
where $\boldsymbol{S}, \R$ and $\boldsymbol{T}$ are matrices\todo{be explicit on dependencies?} for scaling, rotation and translation, respectively. These are provided for reference in Appendix \ref{transformation_matrices}.

Next we map the model into world coordinates using the view matrix $\V$. This involves aligning and positioning the model into the view of the camera. In our case the camera represents the phase center of the sonar, and it is placed in the world origin. The transformation take the form of
\begin{align}
\V &= \mat{T}(x_\text{c}, y_\text{c}, z_\text{c}) \cdot \R_\text{x}(\phi_x, 0, 0) \nonumber\\
&= 
\bmat{
1  &  0  &  0  &  x_\text{c} \\
0  &  1  &  0  &  y_\text{c} \\
0  &  0  &  1  &  z_\text{c} \\
0  &  0  &  0  &  1 \\
} \cdot
\bmat{
1  &  0           &  0           &  0 \\
0  &  \cos\phi_x  &  -\sin\phi_x &  0 \\
0  &  \sin\phi_x  &  \cos\phi_x  &  0 \\
0  &  0           &  0           &  1 \\
},
\end{align}
where $(x_\text{c}, y_\text{c}, z_\text{c})$ is the image center coordinates for along track, cross track, and depth, respectively, and $\Phi_x$ is the elevation angle\todo{is this correct?} of the sonar. It is given as
\begin{align*}
\phi_x &= \arctan\left(\frac{y_\text{mean}}{-z_\text{mean}}\right).
\end{align*}\todo{use absolute sign?}

Finally the relevant part of the scene is projected onto the image plane with a projection matrix $\P$. There are two main types of projection, perspective and orthographic. Perspective projection scales all $x$ and $y$ coordinates inversely with the distance to the objects to create a sense of perspective in the image. Orthographic projection, on the other hand, maps all the scene's vertices to a squared cube where each dimension has the range [-1,1]. This will render all objects equally large whether they are in the distance or up close. The matrix that gives orthographic projection is defined as
\begin{align}
\P = \left(\begin{matrix}
\frac{2}{r-l}  &  0              &  0              &  -\frac{r+l}{r-l} \\
0              &  \frac{2}{t-b}  &  0              &  -\frac{t+b}{t-b} \\
0              &  0              &  \frac{-2}{f-n} &  -\frac{f+n}{f-n} \\
0              &  0              &  0              &  1
\end{matrix}\right)
\end{align}
where $l,r,b$ and $t$ are the left, right, bottom and top boundaries of the projection plane, respectively, and where $n$ and $f$ is the near and far limits of the cube that encompass the scene. These parameters need to be adapted to provide a 2D image that matches the image size, which we specify in along track and cross track coordinates ($x,y$). This is straightforward for left and right since matches the $x$-axis
% They define the window through which we observe our scene.
\begin{align}
l &= x_\text{min}  &  r &= x_\text{max}
\end{align}
For the near and far limits one simply needs to make sure the scene will fit in it
\begin{align}
n &= 0\;\text{m}   &  f &= 300\;\text{m} 
\end{align}
How to set the bottom and top bounaries can be inferred from \Fig{camera_top_bottom}. First we define an origin vector that points to the center of the image
\begin{align}
\vec o &= \left[\frac{x_\text{min}+x_\text{max}}{2}, \frac{y_\text{min}+y_\text{max}}{2}, \frac{z_\text{min}+z_\text{max}}{2} \right]\T,
\end{align}
and the distance from the image origin to its $y$-boundaries is given as
\begin{align}
\Delta y &= \frac{y_\text{max} - y_\text{min}}{2}.
\end{align}
To find the nearest and farthest point in the image that will be part of the projected scene, $\alpha$ and $\beta$, we must consider the seafloor rotation angle around the $x$-axis, $\Phi_x$ 
\begin{align}
\alpha &= \vec 0 - \left[0, \Delta y, \Delta y \times \arctan\Phi_\text{x} \right]
\end{align}

\begin{figure}[t]\centering%
\graphicsAI[drawing,width=\linewidth]{gfx/opengl_part.svg}%
\caption{Setting camera top and bottom boundaries.}\label{camera_top_bottom}%
\end{figure}

@@

\begin{itemize}
\item Kernel tuning.
\item Sharing. PBO. Asynchronous.
\end{itemize}

@@

\begin{itemize}
\item Write to file
\end{itemize}
